On the analytic structure of the linear response susceptibility in open systems

### 2018-03-07 AnalyticStructure_main.tex
update

\begin{abstract}
By performing linear-response (LR)-TDDFT calculation of molecular systems within a highly complete real-space basis set, we analyze how the behavior of the dynamical 
polarizability is influenced by the discretization conditions of the computational treatment.
We show that optical excitations of energy below the LR ionization potential behave as observable, localized quantities, that can be studied with high precision, 
provided an adequate level of completeness. We then present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme. Under this light, we also show that excitation energies above ionization threshold do not exhibit such observable features and \emph{cannot} be 
considered as poles of the dynamical polarizability. This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
unoccupied subspace.
\end{abstract}

@@

In the framework of the linear response formalism one is typically interested in computing the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
{\cal R}_{\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
that represents the perturbation-induced modification of the expectation value of an observable $\op O$\footnote{For a sake of concreteness, we are limiting our considerations 
to observables that do not explicitly depend on time}. Here $\dm'(\omega)$ is the \emph{response density operator} that describes the $\omega$-dependent modification of the density 
operator induced by the perturbation.

\subsection{Basic equation for the linear response formalism for electronic excitations}

The linear response formalism can be effectively applied to the analysis of the electronic excitations of molecular and condensed-matter systems, described by the (density-dependent) 
ground-state hermitian Hamiltonian $\hnot$, coupled to an external perturbation described by the perturbing field $\op\Phi(\omega)$. The $\omega$ structure of $\op\Phi(\omega)$ codifies both 
the intrinsic time dependence of the perturbation and the switching-on procedure, \emph{e.g.} adiabatic or sudden limit protocols.

The response density is expressed as the first-order variation of ground state density
\be\label{rhodef1}
\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
\ee
where $\{\psi_p\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy lower than the Fermi level.
By considering the linear order time evolution of these states in presence of the perturbation we can express the response density in the $\omega$ domain as follows:
\be\lb{rhoPrimeOmegaDef1}
\dm'(\omega) = \ii \commutator{\dmnot}{\op A(\omega)} \;,
\ee
where $\op A(\omega)$ is a sum of two terms, both written as a triple convolution with the retarded/advanced Green functions $\op G_0^{\pm}$ of the unperturbed Hamiltonian:
\be
\op A(\omega) = \frac{1}{2\pi} \int_{-\infty}^\infty  \dd \omega'
\op G_0^+(\omega') \left(\op \Phi(\omega)+\op V'(\omega)\right) \op G_0^-(\omega'-\omega)\;. \nn
\ee
The presence of the second addend is due to the density dependence of the unperturbed Hamiltonian and describes the perturbation induced by the variation of $\hnot$ under the
action of the perturbing field. In the linear response regime this quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel 
(evaluated at $\dm=\dmnot$) times the value of the induced density modification, that is:
\be\label{VpDef1}
\op V'(\omega)= 
\int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
\trace{\dm'(\omega) \ket{\r} \bra{\r'}} \;.
\ee

\subsection{Fluctuation states. A localization argument for the linear response functional}

We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \quad
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. Fluctuation states are, by 
definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
\ee
and consequently the linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee

>>>>

\begin{abstract}
We analyze how the fundamental equations which govern the linear response properties of a quantum system are influenced by the presence of open boundaries.
In particular we focus our attention on linear-response (LR)-TDDFT calculation of molecular systems. We show that optical excitations of energy below the LR ionization potential 
behave as observable, localized quantities, that can be computationally calculated with high precision, provided an adequate level of completeness.
Above this energy threshold, the analytic structure of the susceptibility functional might be separated in two sectors: one can still be described in terms of observable excitation 
with discrete energies, that can be captured within computational setups tailored for localized states. However such excitations are embedded in a continuum of excited states 
that, on the contrary, behave as plane waves and cannot be described efficiently by the same computational techniques. Such excitations do not exhibit observable features and 
\emph{cannot} be considered as poles of the dynamical polarizability. Nonetheless, the contribution of such continuum of excitation reveals to be of paramount importance even 
for expressing LR quantities in the infrared regime like for example static polarizability.
This is particularly evident for small molecules; we show how these considerations might be influenced by the size of molecular system.
We then present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
unoccupied subspace.
\end{abstract}

@@

Linear response can be effectively applied to the analysis of the electronic excitations in molecular and condensed-matter systems. In this framework one consider a system 
described by the Hamiltonian $\hnot$ and assume that the ground-state density $\dmnot$ has been provided. Linear response formalism allows us to evaluate the modification 
of the expectation value of a generic observable induced by a time-dependent perturbation acting on the system. This quantity, written in the frequency domain, can be expressed 
through the evaluation of the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
{\cal R}_{\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where the operator $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} 
under inspection and the \emph{response density operator} $\dm'(\omega)$ codifies the perturbation induced modification of the ground-state density.

The analysis presented in this section is devoted to a critical review of the fundamental equations at the basis of the response density in the linear response framework.    
The aim of this process is to....

%the construction of two different representations of the response density operator...

\subsection{Fluctuation states. A localization argument for the response density}

The response density operator is expressed as the first-order variation of ground state density
\be\label{rhodef1}
\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
\ee
where $\{\ket{\psi_p}\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy $\eps_p$ lower than the Fermi level.
By considering the linear order time evolution of these states in presence of the perturbation we can derive the equation of motion of the response density in the form of 
a quantum Liouville equation \cite{baroni2008}. This procedure suggests us to introduce an explicit representation of the response density in terms of a set of 
$\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, as follows: 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
An analysis of the Liouville equation for the response density written in this fashion evidences that fluctuation states are orthogonal to the occupied eigenstates 
of $\hnot$ and satisfy the following equation of motion:
\be\lb{fluctuationStateEqMotion1}
\left[\bbomega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.
\ee

breve commento su sternheimer e significato lato destro equazione

\vspace{1cm}
Much more interestingly for the main topic of the present discussion, and regardless from the detailed strategy of their solutions, equations \eqref{fluctuationStateEqMotion1} 
allow us to present a \emph{localization argument} for the fluctuation states. We discuss this point for the right vectors, \emph{i.e.} for the first line of \eqref{fluctuationStateEqMotion1}, 
but the same argument holds also for the left ones. We split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term, so that the equation of motion 
for $\ket{f_p(\omega)}$ written in the coordinate representation has the formal structure of an inhomogeneous Helmholtz equation:
\be
\left(k^2_p(\omega)+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;,
\ee
where $k^2_p(\omega) = 2(\omega+\eps_p)$ and we have formally recognized a \emph{source} term of the form:
\be
s(\omega,\r) =  \bra{\r}\op V\ket{f_p(\omega)} + \bra{\r}\op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} 
\ee
The solution of this equation admits an integral representation in the form:
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing, the leading contribution of the fluctuation states
is given by:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k^2_p(\omega)+\ii\eta} |\r|}}{|\r|} \simeq 
e^{-\frac{\eta}{2}|\r|} \frac{e^{\ii \sqrt{k^2_p(\omega)} |\r|}}{|\r|} \;.
\ee
We observe that the presence of the regularization parameter $\eta$ ensures a (slow) asymptotic convergence to zero, nonetheless the qualitative $\r$-dependence of the fluctuation 
states is determined by the sign of $k^2_p(\omega)$. In particular equation \eqref{fluctuationStateIntEq2} describes an exponential damping for $\omega < |\eps_p|$ and an oscillating 
power law behavior for $\omega \geq |\eps_p|$.
This argument holds for all the fluctuation states that contribute to the perturbation-induced expectation value \eqref{LinearResponseFunctDef1}, so we conclude that
the linear response functional is built in terms of localized fluctuation states for $\omega < |\eps_h|$. Instead, in the range $\eps_h|\leq\omega<|\eps_0|$ it contains both
contributions from local and non-local quantities and, finally is composed only by non-local fluctuation states for $\omega\geq|\eps_0|$.

\subsection{Excitation operators. A spectral decomposition representation for the response density}

..........

\subsection{Comments. The analytic structure of the linear response susceptibility}

...........

\subsection{...da togliere.......}

The presence of the second addend is due to the density dependence of the unperturbed Hamiltonian and describes the perturbation induced by the variation of $\hnot$ under the
action of the perturbing field. In the linear response regime this quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel 
(evaluated at $\dm=\dmnot$) times the value of the induced density modification, that is:
\be\label{VpDef1}
\op V'(\omega)= 
\int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
\trace{\dm'(\omega) \ket{\r} \bra{\r'}} \;.
\ee

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% In the framework of the linear response formalism one is typically interested in computing the \emph{linear response functional}:
% \be\lb{LinearResponseFunctDef1}
% {\cal R}_{\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
% \ee
% that represents the perturbation-induced modification of the expectation value of an observable $\op O$\footnote{For a sake of concreteness, we are limiting our considerations 
% to observables that do not explicitly depend on time}. Here $\dm'(\omega)$ is the \emph{response density operator} that describes the $\omega$-dependent modification of the density 
% operator induced by the perturbation.
% 
% \subsection{Basic equations for the linear response formalism for electronic excitations}
% 
% The linear response formalism can be effectively applied to the analysis of the electronic excitations of molecular and condensed-matter systems, described by the (density-dependent) 
% ground-state hermitian Hamiltonian $\hnot$, coupled to an external perturbation described by the perturbing field $\op\Phi(\omega)$. The $\omega$ structure of $\op\Phi(\omega)$ codifies both 
% the intrinsic time dependence of the perturbation and the switching-on procedure, \emph{e.g.} adiabatic or sudden limit protocols. 
% 
% The response density is expressed as the first-order variation of ground state density
% \be\label{rhodef1}
% \dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
% \ee
% where $\{\psi_p\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy lower than the Fermi level.
% By considering the linear order time evolution of these states in presence of the perturbation we can express the response density in the $\omega$ domain as follows:
% \be\lb{rhoPrimeOmegaDef1}
% \dm'(\omega) = \ii \commutator{\dmnot}{\op A(\omega)} \;,
% \ee
% where $\op A(\omega)$ is a sum of two terms, both written as a triple convolution with the retarded/advanced Green functions $\op G_0^{\pm}$ of the unperturbed Hamiltonian:
% \be
% \op A(\omega) = \frac{1}{2\pi} \int_{-\infty}^\infty  \dd \omega'
% \op G_0^+(\omega') \left(\op \Phi(\omega)+\op V'(\omega)\right) \op G_0^-(\omega'-\omega)\;. \nn
% \ee
% The presence of the second addend is due to the density dependence of the unperturbed Hamiltonian and describes the perturbation induced by the variation of $\hnot$ under the
% action of the perturbing field. In the linear response regime this quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel 
% (evaluated at $\dm=\dmnot$) times the value of the induced density modification, that is:
% \be\label{VpDef1}
% \op V'(\omega)= 
% \int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
% \trace{\dm'(\omega) \ket{\r} \bra{\r'}} \;.
% \ee
% 
% \subsection{Fluctuation states. A localization argument for the linear response functional}
% 
% We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
% operator $\op O$, we may write:  
% \be
% \commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
% \ee
% where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
% \be
% \dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
% \ee
% This equation suggests us to define the \emph{fluctuation states}:
% \be\lb{fluctuationStateDef1}
% \ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \quad
% \bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
% \ee
% where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. Fluctuation states are, by 
% definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
% \be
% \brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
% \ee
% The response density written in terms of the fluctuation states reads:
% \be\lb{rhoPrimeFluctuationStateDef1}
% \dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
% \ee
% and consequently the linear response functional can be expressed as:
% \be\lb{LinearResponseFunctDef2}
% {\cal R}_{\op O}(\omega) = 
% \sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
% \ee

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
